Have you forgotten?' Conversion experiences in lists

### 2016-10-24 lists_conversion_experience.rmd
revising manuscript

**How do we approach lists as objects of research?** While lists such as recommendations intersect strongly with other meaning and sense-making practices in everyday life and popular culture, they are also somewhat refractory to discourse  and textual analysis methods. On the one hand, lists are highly associative. They can be interpreted or decoded semiotically, but they also exhibit great variations in textuality that multiply semiotic readings. On the other hand, as operational inscriptions, they can be treated ethnomethodologically, as the production of ('the order of social settings through ...  shared sense making practices.')[https://en.wikipedia.org/wiki/Ethnomethodology]

@@

To this end, I will be seeking to simulate the workings of the recommender system by generating queries, and exploring results from  a very large supermarket chain database.

I respond to the presentation in two ways. First, by recounting, contextualising and commenting on some of the main steps in making the 'Have your forgotten?' list  as a example of the sense-making done by data scientists and customer analytics teams in many places. Many of these steps were mentioned only briefly in the presentation, and they are largely taken for granted as part of predictive analytic practice.  Second, by trying to step into the list-making practices  as an occasion to intervene and participate in the ontological conversion practices they engage in. This is no mere methodological problem if, as Mark Hansen has recently suggested,

@@

Shreena Patel works as a data scientist for DunnHumby, a well-known [ customer science company ](https://www.dunnhumby.com/)  focusing on online grocery shopping at Tesco. I found her presentation on shopping lists to be  practical and evocative since it was focused quite a lot on implementation of recommendations, and symptomatic in various ways.

![Tesco ClubCard](figure/tesco-clubcard.jpg )

@@

While Tesco has been very successful in data-mining its customers, the online shopping situation is a bit different. It is no longer organised around campaigns involving special offers or redemption of points, but on continuous up-selling and cross-selling.

- the problems of scale: Tesco is like Amazon or Netflix in this respect: too many people and too many products

>Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet. (https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market/)[https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market/]

@@

- online commerce -- transforming shopping; 
- personalization and recommendation as the killer apps for social media as a commercial process, but also in government etc.

How would I characterise this shift? Use it to emphasize why attention to the algorithm is not enough -- it is the model that counts
A little history of recommender systems -- uses Varian 1997 for a start. But has no real reference list. 
Also can use Schafer -- back to Association Rule Mining from early 1990s

When commentators first began talking about recommender systems in the 1990s, they highlighted the shift from demographic-driven market research techniques to personalized recommendations. For instance, writing in 1997, Paul Resnick and Hal Varian (at that time Dean of Information Sciences at UC Berkeley, but somehow now Chief Economist at Google), made much of this difference [@Resnick_1997].

@@

@@

Patel described her work at DunnHumby in terms of a shift in models from a 'rules-based list' to a 'relevancy model.' This was a significant change in how the 'Have You Forgotten' list was constructed from one kind of model to another. For me, the shift in models matters since it amounts to a much more probabilistic structuring of lists.

I assume that the rules-based model derived from the early 1990s work on the `apriori` algorithm by Rakesh Agrawal and Ramakrishnan Srikanti at IBM Research Alameda [@Agrawal_1994]. The `apriori` algorithm finds sets of items that commonly occur in transactions. These commonly occurring sets are expressed as association rules. For instance, using the `Groceries` dataset in the `R` package `arules,` we can readily see how Walmart might have discovered  -- but didn't -- the fabled association between diapers and beer.  [http://www.dssresources.com/newsletters/66.php]
The `apriori` algorithm is widely used in 'market basket analysis' and attempts to re-structure shopping lists in terms of associations between items that are frequently bought together.

```{r arules, cache=TRUE, echo=TRUE, fig.margin=TRUE}
library(knitr)
library(arules)
library(arulesViz)
library(datasets)
data(Groceries)
itemFrequencyPlot(Groceries,topN=20,type="absolute")
rules <- apriori(Groceries, parameter = list(supp = 0.001, conf = 0.8))

@@

Badiou in his account of multiples highlight the contrast between two different ways of grasping a multiple -- in terms of the 'numerical one' and the 'vital One' [@Badiou_2004, 72], between the 'closed' and the 'open' respectively.

The `apriori` algorithm, might be seen as an attempt to deal with the excess of things in a Tesco supermarket, and the lack of interest or engagement that people have with them by identifying subsets whose membership is stronger by virtue of the other sets of associations that people bring with them.   
Association is stronger when it derives from an excess of different ways of being in sets.

Badiou's argument is that being part of a set differs from belonging to a set.
An open set arises from when we dwell within a set by 'taking the intersection of two elements.' What might this mean in terms of ordering?

If we take the itemset from the `Groceries` dataset [@Hahsler_2006], and then run it through the Tesco online groceries site, each of the 169 items in the dataset yields dozens and sometimes thousands of results from Tesco online.

@@

Each of the items in the `Groceries` dataset proliferates in the Tesco site, into a new list of branded variations. The 169 items of the `Groceries` dataset, itself from a local German or Austrian supermarket [@REF_TBA],  expand into roughly `r total` possible items.

@@

Patel observed that the 'Have you forgotten' lists matter to people, and we can see why in the light of this proliferation of things in the list. A `r total` list of items is quite hard to list let alone read. 
Patel said that some people log on and go straight to the checkout, allowing the recommender system to supply them with the suggestions that they then move to the shopping basket.
This is the online equivalent of using an old shopping list and only adding new items when needed. 
It is also a strategy to avoid having to view too many lists, since the only way Tesco can show people what they might buy is by showing them lists of things (e.g. chicken products), but these lists are often too long to work with. For instance, Tesco has more than 1000 chicken products on its site.   
People inhabit lists and work with them no matter how they are made; thus vindicating ethnomethodologists assertion that social order is the product of ongoing sense-making practices. 
But it also suggests a transformation in the way that lists exist in the world. They are less the product of a discrete agent than an effect of an environment teeming with possible combinations and complementaries, a surplus of possible purchases that the supermarket needs to continually maintain but also filter in order to make it possible for something to happen.

## 5. Data: 52 weeks of data for each customer: how is order constructed -- from history?

Patel described the data they on grocery purchases. 'We have 52 weeks of purchase data for each customer.' 
We know that the construction of lists has been transformed by recommender systems.
These transformations have been widely discussed in the context of big data, where questions of who has how much data are usually most prominent.
In terms of the ways that lists order the world, the amount of data is not so amount as the way in which things in the world take on a list-like existence. 
I want to dwell on this idea for a while. We have seen two different ways in which lists might be constructed in a recommender system. 
The first uses history and existence of previous lists, including the customer's own list, to populate a new list. 
The second uses many shopping lists to identify associations between things and then uses the strongest associations to recommend additions to a current list. 
The 'relevancy model,' however, tries to do something different. 
The problem that the proliferation of things on the shelves of supermarket or grocery warehouse confronts is partly combinatorial.
With  `r total` products (actually Patel mentions 200,000 products) can be combined in truly astronomical numbers of combinations.
If a typical shopping list has 20 items, then there are `r format(choose(total, 20), 2)` possible combinations.
Some of this vast number of possible shopping lists is highly unlikely in terms of shopping lists and histories of shopping, but potentially highly interesting and value as recommendation.

>>>>

## How do we approach lists as objects of research?

While lists such as recommendations intersect strongly with other meaning and sense-making practices in everyday life and popular culture, they are also somewhat refractory to discourse  and textual analysis methods. On the one hand, lists are highly associative. They can be interpreted or decoded semiotically, but they also exhibit great variations in textuality that multiply semiotic readings. On the other hand, as operational inscriptions, they can be treated ethnomethodologically, as the production of ('the order of social settings through ...  shared sense making practices.')[https://en.wikipedia.org/wiki/Ethnomethodology]

@@

This is no mere methodological problem if, as Mark Hansen has recently suggested,

@@

Shreena Patel works as a data scientist for DunnHumby, a well-known [ customer science company ](https://www.dunnhumby.com/)  focusing on online grocery shopping at Tesco. I found her presentation on shopping lists to be  practical and evocative since it was focused quite a lot on implementation of recommendations, and symptomatic in various ways. It was filled with graphs, numbers, and tables concerning different aspects of the 'have you forgotten?' list. Understandably, this was not presented in a language of ontological transformation.

To this end, I will be seeking to reconstruct some of the workings of the recommender system by generating queries, and exploring results from  a very large supermarket chain database.

I respond to the presentation in two ways. First, by recounting, contextualising and commenting on some of the main steps in making the 'Have your forgotten?' list  as a example of the sense-making done by data scientists and customer analytics teams in many places. Many of these steps were mentioned only briefly in the presentation, and they are largely taken for granted as part of predictive analytic practice.  Second, by trying to step into the list-making practices  as an occasion to intervene and participate in the ontological conversion practices they engage in.

!['Every little bit helps':Tesco ClubCard](figure/tesco-clubcard.jpg )

@@

While Tesco has been very successful in data-mining its customers,  and perhaps became the UK's biggest supermarket through data-mining in the 1990s, the online shopping situation is a bit different. It is no longer organised around campaigns involving special offers or redemption of points, but on continuous up-selling and cross-selling to customers who might be anywhere, but especially are not in the supermarket itself. .

>Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet. [ https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market/ ]( https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market/ )

@@

How should characterise this shift from ClubCard membership data mining to online grocery markets?  Patel described her work at DunnHumby in terms of a shift in models from a 'rules-based list' to a 'relevancy model.' This was a significant change in how the 'Have You Forgotten' list was constructed from one kind of model to another. For me, this shift is a part of broader re-organisation of prediction in which existing predictive practices are being re-distributed and intensified in particular ways. The shift in models matters since it amounts to a much more probabilistic structuring of lists.

When commentators first began talking about recommender systems in the 1990s, they highlighted the shift from demographic-driven market research or data mining techniques to personalized recommendations. For instance, writing in 1997 in a special section of the _Communications of the ACM_ on recommender systems, Paul Resnick and Hal Varian (at that time Dean of Information Sciences at UC Berkeley, but somehow now Chief Economist at Google), made much of this difference [@Resnick_1997]. They emphasise the need for a term to distinguish the emerging practices from data mining. [^44]

[^44]: Also can use Schafer -- back to Association Rule Mining from early 1990s

@@

> In everyday life, we rely on recommendations from other people. ... Recommender systems augment this natural social process. In a typical recommender  system, people provide recommendations as inputs, which the system then aggregates and directs to appropriate recipients. In some cases the primary transformation is in the aggregation; in others the system’s value lies in its ability to make good matches between the recommenders and those seeking recommendations.[@Resnick_1997,56]

Writing near the advent of early web-based e-Commerce, Resnick and Varian (now Chief Economist at Google) imagined recommender systems as automations of the 'collaborative filtering' that people do when they write and read reviews of products and services (for instance, on the travel website, TripAdvisor; [@REF_Scott_TBA]

@@

We can see something of this shift from collaborative filtering in the shift that Patel describes as moving from 'a rules-based' model to personal relevance. I assume that the rules-based model derived from the early 1990s work on the `apriori` algorithm by Rakesh Agrawal and Ramakrishnan Srikanti at IBM Research Alameda [@Agrawal_1994]. The `apriori` algorithm finds sets of items that commonly occur in transactions. It was listed as one of the top ten data mining algorithms in a survey conducted amongst data miners [@TBA_2007].  These commonly occurring sets are expressed as association rules. For instance, using the `Groceries` dataset in the `R` package `arules,` we can readily see how Walmart might have discovered  -- but didn't -- the fabled association between diapers and beer.  [http://www.dssresources.com/newsletters/66.php]

The `apriori` algorithm is widely used in 'market basket analysis' and attempts to re-structure shopping lists in terms of associations between items that are frequently bought together.

```{r arules, cache=TRUE, echo=TRUE, fig.margin=TRUE}
library(knitr)
library(arules)
library(arulesViz)
library(datasets)
data(Groceries)
itemFrequencyPlot(Groceries,topN=10,col=rainbow(10), ntype="absolute",  cex.axis = 0.55)
rules <- apriori(Groceries, parameter = list(supp = 0.001, conf = 0.8))

@@

Badiou in his account of multiples highlights the contrast between two different ways of grasping a multiple -- in terms of the 'numerical one' and the 'vital One' [@Badiou_2004, 72], between the 'closed' and the 'open' respectively. Note that Badiou does not explicitly oppose the closed and open. While Hansen tends to see predictive analytics as the closed-formed of the micrological propensities of things,  Badiou refers to a 'joint effectuation.'

What would be Badiou's take on a grocery list as augmented by a recommender system? The `apriori` algorithm, might be seen as an attempt to grasp the excess of things in a Tesco supermarket by identifying subsets whose membership is stronger by virtue of the other sets of associations that people bring with them.   
Association is stronger when it derives from an excess of different ways of being in sets.

Badiou's argument is that being part of a set differs from belonging to a set. An open set arises from when we dwell within a set by 'taking the intersection of two elements.' What might this mean in terms of ordering carried out with specific grocery items in a supermarket?

If we take a typical grocery dataset used in data-mining modelling (e.g. from the `Groceries` dataset [@Hahsler_2006a] acquired from 'local German supermarket'), and then paste it into the 'shopping list' box on on the Tesco grocery website, each of the 169 items in the dataset yields dozens and sometimes thousands of results from Tesco online.

@@

Each of the items in the `Groceries` dataset proliferates in the Tesco site, into a new list of branded variations. The 169 items of Hahsler `Groceries` dataset  expand into roughly `r total` possible items.  This is, I would suggest, a form of open. The multiplicity grasped at the `Grocery` dataset becomes more open in this setting, where 'open' complements any possible closure with new intersections. Rather than seeing predictive analytics as a closed form of an open world, I am suggesting that we might begin to locate it in the joint effectuations or intersections within models.

@@

Patel said that some people log on and go straight to the checkout, allowing the recommender system to supply them with the suggestions that they then move to the shopping basket. Patel observed that the 'Have you forgotten' lists matter to people, and we can see why in the light of this proliferation of things in the list. A `r total` list of items is quite hard to list let alone read. This is the online equivalent of using an old shopping list and only adding new items when needed. 
It is also a strategy to avoid having to view too many lists, since the only way Tesco can show people what they might buy is by showing them lists of things (e.g. chicken products), but these lists are often too long to work with. For instance, if Tesco has more than 1000 chicken products on its site, how can a customer even look through the list of chicken products?

People inhabit lists and work with them no matter how they are made; thus vindicating ethnomethodologists assertion that social order is the product of ongoing sense-making practices. [^45] But it also suggests a transformation in the way that lists exist in the world. They are less the inscriptions of a discrete agent than an effect of an environment teeming with possible combinations and complementaries, a surplus of possible purchases that the supermarket needs to continually maintain but also filter in order to make it possible for something to happen.

[^45]: What have ethnomethodologists and others actually said about lists?

## 5. 'We have 52 weeks of data for each customer': how is order constructed -- from history?

We know that the construction of lists has been transformed by recommender systems.
Patel described the data that Tesco keeps on grocery purchases: 'We have 52 weeks of purchase data for each customer.' 
These transformations have been widely discussed in the context of big data, where questions of who has how much data are usually most prominent (see [@Kitchin_2014]).
In terms of the ways that lists order the world, the amount of data is not perhaps not so important as the way in which list-making is re-distributed in the world. 
I want to dwell on this idea for a while.

HERE

We have seen two different ways in which lists might be constructed in a recommender system. 
The first uses history and existence of previous lists, including the customer's own list, to populate a new list. 
The second uses many shopping lists to identify associations between things and then uses the strongest associations to recommend additions to a current list. 
The 'relevancy model,' however, tries to do something different. 
The problem that the proliferation of things on the shelves of supermarket or grocery warehouse confronts is partly combinatorial.
With  `r total` products (actually Patel mentions 200,000 products) can be combined in truly astronomical numbers of combinations.
If a typical shopping list has 20 items, then there are `r format(choose(total, 20), 2)` possible combinations.
Some of this vast number of possible shopping lists is highly unlikely in terms of shopping lists and histories of shopping, but potentially highly interesting and value as recommendation.
